---
title: "Red Wine Quality"
author: Ahmed Eweed
date: June 9, 2018
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---

# Introduction  

This is one of two datasets involving red and white wine samples,  
Created by: Paulo Cortez (Univ. Minho), Antonio Cerdeira, Fernando Almeida,  
Telmo Matos and Jose Reis (CVRVV) @ 2009.  
It involves 1599 observations about 11 chemical features and  
an output result (quality) of the Portuguese "Vinho Verde" wine.  
Wine trade is huge around the world from planting grapes, fermentation and  
bottling of the finished liquid.  
I wondered which factor affects a wine taste and quality the most?  
Is that sugar, alcohol percentage or maybe it's one of the acids  
or other chemicals added to the wine through the production process.  
We will find out if my assumptions were right!  

Citation: P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis.  
Modeling wine preferences by data mining from physicochemical properties.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
#loading libraries
library(ggplot2)
# Loading Data, rw for 'red wine'
rw <- read.csv("wineQualityReds.csv")
```
I will explore the distribution of the features across the dataset  
to see if there is any abnormal behaviours or misleading data
# Univariate Plots Section
```{r echo=FALSE, Univariate_Plots, message=FALSE, warning=FALSE}
#exploring dataset strcture 
str(rw)
```

we can see 13 variables of 11 features and a result score (quality) plus a counter (X)

```{r echo=FALSE, message=FALSE, warning=FALSE}
#quality bar chart
ggplot(aes(quality), data = rw)+
  geom_bar(binwidth = 1/10)+
  xlab("Quality")+
  ylab("Number of wines")+
  ggtitle("Red Wine Quality distribution")

summary(rw$quality)
```
red Wine quality is normally distributed with mean of about 5.5 

```{r echo=FALSE, message=FALSE, warning=FALSE}
#fixed acidity bar chart
ggplot(aes(fixed.acidity), data = rw)+
  geom_bar(binwidth = 1/10)+
  xlab("Fixed Acidity (g/dm^3)")+
  ylab("Number of Wines")+
  ggtitle("Fixed Acidity distribuiton")

summary(rw$fixed.acidity)
```
Fixed Acidity is skewed to the right with mean of 8

```{r echo=FALSE, message=FALSE, warning=FALSE}
#volatile acidity bar chart
ggplot(aes(volatile.acidity), data = rw)+
  geom_bar(binwidth = 1/100)+
  xlab("Volatile Acidity (g/dm^3)")+
  ylab("Number of Wines")+
  ggtitle("Volatile Acidity distribution")

summary(rw$volatile.acidity)
```
Volatile Acidity is skewed to the right with a mean of 0.5

```{r echo=FALSE, message=FALSE, warning=FALSE}
#citric acid bar chart
ggplot(aes(citric.acid), data = rw)+
  geom_bar(binwidth = 1/50)+
  xlab("Citric Acid (g/dm^3)")+
  ylab("Number of Wines")+
  ggtitle("Citric Acid distribution")

summary(rw$citric.acid)
```
Citric Acid is not normally distributed!

```{r echo=FALSE, message=FALSE, warning=FALSE}
#residual sugar boxplot
ggplot(aes(residual.sugar), data = rw)+
  geom_bar(binwidth = 1/80)+
  ylab("Number of wines")+
  xlab("Residual Sugar (g/dm^3)")+
  scale_x_log10(breaks =seq(0.85, 15, 3),
                limits =c(0.85, 15))+
  ggtitle("Residual Sugar distribution")

summary(rw$residual.sugar)
```
Residual Sugar is left skewed with many outliers on the right 
and more the 75% of the data are concentrated under 2.6  
this distribution is suspicious!

```{r echo=FALSE, message=FALSE, warning=FALSE}
#chlorides bar chart
ggplot(aes(chlorides), data = rw)+
  geom_bar(binwidth = 1/200)+
  xlab("Chlorides (g/dm^3)")+
  ylab("Number of wines")+
  scale_y_log10()+
  ggtitle("Chlorides distribution")

summary(rw$chlorides)
```
Chlorides are not normally distributed with mean of about 0.1 and have many outliers

```{r echo=FALSE, message=FALSE, warning=FALSE}
#free sulfur dioxide bar chart
ggplot(aes(free.sulfur.dioxide), data = rw)+
  geom_bar(binwidth = 1)+
  xlab("Free Sulfur Dioxide (mg/dm^3)")+
  ylab("Number of Wines")+
  ggtitle("Free Sulfur Dioxide distribution")

summary(rw$free.sulfur.dioxide)
```
Free Sulfur Dioxide is skewed to the right with mean at 15

```{r echo=FALSE, message=FALSE, warning=FALSE}
#total sulfur dioxide bar chart
ggplot(aes(total.sulfur.dioxide), data = rw)+
  geom_bar(binwidth = 5)+
  xlab("Total Sulfur Dioxide (mg/dm^3)")+
  ylab("Number of Wines")+
  ggtitle("Total Sulfur Dioxide distribution")

summary(rw$total.sulfur.dioxide)
```
Total Sulfur Dioxide is skewed to the right with mean of 46 

```{r echo=FALSE, message=FALSE, warning=FALSE}
#density bar chart
ggplot(aes(density), data = rw)+
  geom_bar(binwidth = 1/700)+
  xlab("Density (g/cm^3)")+
  ylab("Number of Wines")+
  ggtitle("Density distribution")

summary(rw$density)
```
Density is normally distributed with mean of 0.99 

```{r echo=FALSE, message=FALSE, warning=FALSE}
#pH bar chart
ggplot(aes(pH), data = rw)+
  geom_bar(binwidth = 1/10)+
  xlab("pH")+
  ylab("Number of Wines")+
  ggtitle("pH distribution")

summary(rw$pH)
```
pH is normally distributed with mean of 3 

```{r echo=FALSE, message=FALSE, warning=FALSE}
#sulphates bar chart
ggplot(aes(sulphates), data = rw)+
  geom_bar(binwidth = 1/10)+
  xlab("Sulphates (g/dm^3)")+
  ylab("Number of Wines")+
  ggtitle("sulphates distribution")

summary(rw$sulphates)
```
sulphates is skewed to the right with mean of 0.6

```{r echo=FALSE, message=FALSE, warning=FALSE}
#alcohol bar chart
ggplot(aes(alcohol), data = rw)+
  geom_bar(binwidth = 1/5)+
  xlab("ALcohol (%)")+
  ylab("Number of Wines")+
  ggtitle("alcohol distribution")

summary(rw$alcohol)
```
alcohol is skewed to the right with mean of 10  

# Univariate Analysis

### What is the structure of your dataset?
We have 1599 observations with 13 variables,
a counter (X), the result score (quality) and 11 chemical features as follows:
(fixed acidity, volatile acidity, citric acid, residual sugar, chlorides,
free sulfur dioxide, total sulfur dioxide, density, pH, sulphates, alcohol)

### What is/are the main feature(s) of interest in your dataset?

I found citric acid and volatile acidity interesting 

### What other features in the dataset do you think
### will help support your investigation into your feature(s) of interest?

I think alcohol and sulphates will be interesting too

### Did you create any new variables from existing variables in the dataset?

no i didn't

### Of the features you investigated, were there any unusual distributions?  
### Did you perform any operations on the data to tidy, adjust,
### or change the form of the data?  
### If so, why did you do this?

citric acid is not normally distibuted
also Residual sugar is suspicious with alot of outlier data


yes, Chlorides and residual sugar was log transformed to avoid outlier presence  

# Bivariate Plots Section
```{r echo=FALSE, Bivariate_Plots, message=FALSE, warning=FALSE}
# scatter of quality and volatile acidity
ggplot(aes(x = factor(quality), y = volatile.acidity), data = rw)+
  geom_boxplot()+
  geom_jitter(alpha = 1/5)+
  xlab("Quality")+
  ylab("Volatile Acidity (g/dm^3)")+
  ggtitle("Boxplot of Volatile Acidity across quality levels")

# correlation between quality and volatile acidity
cor.test(rw$volatile.acidity, rw$quality, method = "pearson")
```
a weak negative correaltion (-0.39) exists between Quality and Volatile Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE}
# scatter of quality and citric acid
ggplot(aes(x = factor(quality), y = citric.acid), data = rw)+
  geom_boxplot()+
  geom_jitter(alpha = 1/5)+
  xlab("Quality")+
  ylab("Citric Acid (g/dm^3)")+
  ggtitle("Boxplot of Citric Acid across quality levels")

# correlation between quality and citric acid
cor.test(rw$citric.acid, rw$quality, method = "pearson")
```
a weak positive correaltion (0.22) exists between Quality and Citric Acid

```{r echo=FALSE, message=FALSE, warning=FALSE}
# scatter between quality and sulphates
ggplot(aes(x = factor(quality), y = sulphates), data = rw)+
  geom_boxplot()+
  geom_jitter(alpha = 1/5)+
  xlab("Quality")+
  ylab("Sulphates (g/dm^3)")+
  ggtitle("Boxplot of Sulphates across quality levels")
# correaltion between quality and sulphates
cor.test(rw$sulphates, rw$quality, method = "pearson")
```
a weak positive correaltion (0.25) exists between Quality and Sulphates  
many outliers exist too

```{r echo=FALSE, message=FALSE, warning=FALSE}
# scatter between quality and alcohol
ggplot(aes(x = factor(quality), y = alcohol), data = rw)+
  geom_boxplot()+
  geom_jitter(alpha = 1/5)+
  xlab("Quality")+
  ylab("Alcohol (%)")+
  ggtitle("Boxplot of Alcohol across quality levels")
# correalation between quality and alcohol
cor.test(rw$alcohol, rw$quality, method = "pearson")
```
a strong positive correaltion (0.47) exists between Quality and Alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE}
# scatter between alcohol and density
ggplot(aes(y = alcohol, x = density, color = factor(quality)), data = rw)+
  geom_jitter()+
  coord_cartesian(xlim = c(0.985, 1.005), ylim = c(5, 15))+
  xlab("Density (g/cm^3)")+
  ylab("Alcohol (%)")+
  scale_color_brewer(type = 'seq')+
  ggtitle("Relaionship of density Vs Alcohol across quality levels")+
  theme_dark()
# correlation between citric acid and volatie acidity
cor.test(rw$alcohol, rw$density, method = "pearson")
```
a strong negative correaltion (-0.49) exists  
between Density and Alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE}
# scatter between alcohol and sulphates
ggplot(aes(y = alcohol, x = sulphates, color = factor(quality)), data = rw)+
  geom_jitter()+
  coord_cartesian(xlim = c(0.33, 2.0), ylim = c(5, 15))+
  xlab("Sulphates (g/dm^3)")+
  ylab("Alcohol (%)")+
  scale_color_brewer(type = 'seq', guide = guide_legend(title = "Quality levels"))+
  ggtitle("Relaionship of Sulphates Vs Alcohol across quality levels")+
  theme_dark()


#correlation between sulphates and alcohol
cor.test(rw$alcohol, rw$sulphates, method = "pearson")
```

there is almost no relationship between Alcohol and Sulphates

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation.
### How did the feature(s) of interest vary with other features in the dataset?

there is negative relationship between quality and volatile acidity
also volatile acidity and citric acid have negative relationship

there is positive relationship between quality and citric acid

### Did you observe any interesting relationships
### between the other features (not the main feature(s) of interest)?

there are positive relationships between:
quality and sulphates, quality and alcohol and sulphates and alcohol

### What was the strongest relationship you found?

the negative relationship between volatile acidity and citric acid
was the strongest so far (-55%)

# Multivariate Plots Section

```{r echo=FALSE, Multivariate_Plots}
ggplot(aes(x = citric.acid, y = volatile.acidity),
       data = rw)+
  geom_jitter(aes(color = factor(quality)))+
  scale_x_continuous()+
  ylab("Volatile Acidity (g/dm^3)")+
  xlab("Citric Acid (g/dm^3)")+
  ggtitle("Quality based on Voaltile Acidity and Citric acid")+
  scale_color_brewer(type = 'seq', guide = guide_legend(title = "Quality levels"))+
  theme_dark()
```

low levels of citric and volatile acids relates to higher quality rates

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = volatile.acidity, y = alcohol), data = rw)+
  geom_jitter(aes(color = factor(quality)))+
  xlab("Volatile Acidity (g/dm^3)")+
  ylab("Alcohol (%)")+
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(title = 'Quality levels'))+
  ggtitle("Alcohol and Volatile Acidity with quality levels")+
  theme_dark()

```


higher Alcohol content and moderate volatile acidity level relates to higher quality rates


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the investigation.
### Were there features that strengthened each other in terms of looking at your feature(s) of interest?

Volatile acidity and Citric Acid has a negaive relationship with each other 
and the less the Volatile acid and the more Citric Acid in the wine,  
the higher rate it takes

### Were there any interesting or surprising interactions between features?
Residual sugar had no effect on the quality

although it should be a major player in rating wines

### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.

no I didn't
------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}

ggplot(aes(quality), data = rw)+
  geom_bar(aes(fill = factor(quality)))+
  xlab("Quality")+
  ylab("Number of wines")
         

```

### Description One

red wine quality is normally distributed with almost no wine rated 0,1,2 and 9  
wine rate(worst----> best):  
3 - 4 - 5 - 6 - 7 - 8

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(y = alcohol, x = density, color = factor(quality)), data = rw)+
  geom_jitter()+
  coord_cartesian(xlim = c(0.985, 1.005), ylim = c(5, 15))+
  xlab("Density (g/cm^3)")+
  ylab("Alcohol (%)")+
  scale_color_brewer(type = 'seq')+
  ggtitle("Relaionship of density Vs Alcohol across quality levels")+
  theme_dark()

```

### Description Two

higher quality is related to denser wines and more alcohol content ones  

### Plot Three
```{r echo=FALSE, Plot_Three, message= FALSE}

ggplot(aes(y = alcohol, x = sulphates, color = factor(quality)), data = rw)+
  geom_jitter()+
  coord_cartesian(xlim = c(0.33, 2.0), ylim = c(5, 15))+
  xlab("Sulphates (g/dm^3)")+
  ylab("Alcohol (%)")+
  scale_color_brewer(type = 'seq')+
  ggtitle("Relaionship of Sulphates Vs Alcohol across quality levels")+
  theme_dark()

```

### Description Three

sulphates between 0.6 and 0.82 g/dm^3 and alcohol between 10 and 14 percent  
has the highest quality rate

------

# Reflection

Red Wine has very interesting quality factors 
I had much fun exploring this dataset and discovering new relations between features  
it was hard to pick the right features as all features are almost chemical  
but, with some research and collecting data about wine production,  
I managed to detect the right ones, or so I thought!  
unfortunatelly, we should have more samples to make the best prediction of them  
As Residual sugar surprisingly was ineffective in determining the quality of wine 
I think more features are necessary to be collected to determine wine quality 
not only chemical ones, but also grape kind and conditions of storing
It was an interesting experience as well.
